### 2022-01-19 plos_latex_template.tex
edits to design rules, esp. Rules 4, 5 ~ @jananiravi

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract (optional from what I've seen)}

% % Batool: I didn't change anything in the abstract - you can check the history - I copied the abstract and pasted, then the colour change for the abstract but it contains all text and suggestions/edits by Janani - my apologies (didn't commit the changes).  
The authors of this article recently organized useR! 2021, the annual global R conference, attracting a broad range of participants from academia, industry, government, and the non-profit sector. To embrace the constraints posed by the pandemic, this year we hosted a 100\% virtual conference. We worked on building a high-quality virtual and explicitly global conference by creating a kind, inclusive, accessible, and welcoming environment for everyone. 
This article outlines our most important lessons learned abstracted as '10 simple rules to host an inclusive conference'. These rules apply equally to academic, industry, or mixed conferences; while inspired by a global experience, these rules seamlessly translate to regional events. These rules were implemented carefully and intentionally during a virtual conference, but are designed to serve as useful guidelines for in-person and hybrid events even after the COVID pandemic.

@@

% Rocío: Main message of the paragraph: Conferences are great, but for some
Conferences, from the Latin \textit{conferre} `to bring together', are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions.
The larger the conference, the larger the opportunities to network and learn from the cohort.
However, opportunities for participating in conferences are not equal for all. 
Many academic and tech conferences have become exclusive spaces for privileged groups, \textit{i.e.}, with unearned advantages given by society to some but not all (\textit{e.g.}, white, male, from a rich country, English-native speaker, with no physical disabilities) \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}.
% However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields.

This article proposes 10 simple rules to host inclusive conferences.
They are organized in three sections: \textbf{the pillars}, fundamental rules with key elements to conceive the work on diversity and inclusion in any conference (embracing diversity, creating a welcoming environment, and having a diverse organizing team); \textbf{design rules}, concerning diverse elements in the design of the conference and operational tasks (counteracting bias in the conference program, embracing a strong online component, achieving accessibility for people with disabilities, embracing language diversity, enforcing inclusive communication strategies, and allocating financial resources to support inclusive practices); and a \textbf{process rule} to make the conference part of a long-term commitment to inclusion. 
These practices stem from the experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. useR! conferences are non-profit events organized by volunteers from the R community and managed by the R Foundation.
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many persons as possible.

The authors of this article include the global coordinators of the conference, members of the diversity and inclusion team, and people who contributed to other areas of the conference. 
Therefore, the rules result from our lessons learned from diverse perspectives and expertise (\textit{i.e.}, our different backgrounds, profiles, and roles in the conference). 
Even though these rules were designed based on our experience with a global event, they can be easily transferred or scaled to events with regional or local scopes.
These rules will greatly benefit people who are part of meeting committees that oversee the site/location selection process, or that coordinates with the local organizers of conferences.
The rules will also help local/virtual organizers who intend to put inclusion at the core of their conference right from the conception and planning phases. We have designed them for wide adoption to enable and ensure that the scientific and technical conferences during and post-pandemic are as accessible and inclusive as they can be.

@@

\begin{figure}[!h]
\centering
%sara: without success i'm trying to use pdfcomment package and pdftooltip to make the alt-text. any help?
%ast: R/ Unfortunately, support for adding accessibility metadata to LaTeX documents is limited. You will probably need to use Adobe Acrobat to add missing accessibility metadata to your PDF file.https://chi2021.acm.org/for-authors/presenting/papers/guide-to-an-accessible-submission#latex
\pdftooltip{
\includegraphics[width=\textwidth]{figs/rules.png}}{
%sara: suggestion for alt text:
%Diagram of how the 10 rules are organized into three groups: grounding rules (rules 1, 2, and 10), people-related rules (rules 3 and 4), and design rules (rules 5 to 9). The diagram has five rows. Each line contains rectangular boxes with the rule number and a short title for each rule. Each rule box is colored based on their group (grounding rules are grey, people-related rules are green and design rules are pink). First row is for the grounding rules 1. Multidimensional diversity and 2. Safe and inclusive environment in grey boxes. Second row is for people-related rules 3. Organizing team and 4. Unbias in green boxes. Third row is for the design rules 5. Online component, 6. Accessibility for disabilities, 7. Language inclusivity in pink boxes. The fourth row is for the design rules 8. Communication and 9. Financial support and budget in pink boxes. The last row is for the grounding rule 10. Diversity and inclusion as a process in grey boxes.
}
\caption{Diagram of the rules organized in three groups: the pillars (rules 1, 2, and 3), design rules (rules 4 to 9), and the process (rule 10).}
\label{fig:diagram}
\end{figure}

@@

To organize a more inclusive conference, it is necessary to identify upfront which groups of people are being excluded, the barriers that keep them from active partaking, and targeted strategies to overcome them. 
\textbf{Rule 1} is about embracing all dimensions of diversity and prioritizing the participation of groups that are often marginalized.
However, increasing diversity alone is not enough. \textbf{Rule 2} focuses on how to create a safe and welcoming environment for all the conference attendees. 
Inclusion should start from the organizing team. \textbf{Rule 3} highlights the importance of starting with an inclusive and diverse organizing team, and provides tips on work dynamics.

@@

% Rocío: Main idea: what is diversity and why does it matter? (inequalities)
Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few.
Human diversity should be celebrated and respected in every way. 
Unfortunately, we live in a world with implicit hierarchies along these axes. 
Some statuses (\textit{e.g.}, cisgender, white, male, from North America or Western Europe) hold the privilege of being the defaults around which all systems -- including conferences -- are consciously and unconsciously built. 
People outside the dominant groups suffer different forms of oppression, like sexism, racism, ableism, homophobia, and transphobia; a few of them have to deal with several of these on a daily basis.
% While no isolated initiative can drastically alter conference outcomes, building a more diverse and inclusive conference starts by recognizing that these inequalities have systematically excluded whole groups of people from academia and scientific and professional circles. 
Therefore, recognizing this systematic exclusion, particularly in our fields and in our scientific or professional communities, is paramount to inclusion, and will help identify the most marginalized and discriminated subgroups \cite{timperleyHeMoanaPukepuke2020}.
% These are the groups that warrant deliberate efforts that ensure fair inclusion.

\emoji{bulb} \textbf{Practical tips:} As a starting point, the meeting organizers could identify which groups warrant deliberate efforts to ensure fair inclusion, and determine the contributing factors (\textit{e.g.}, explicit discrimination, participation fees, etc.).
This analysis should guide the vision of diversity and inclusion for your conference and help design strategies to achieve it. 
Express this vision for your conference in a diversity statement. 
Next, use this vision and the set strategies to define measurable indicators and goals of diversity and inclusion (\textit{e.g.}, a gender distribution of your speakers that is representative of the general population, presence of racialized people --especially Black people-- among the head organizers, speakers, and attendees, or participation from key geographic regions). 
These numbers should guide you and help achieve your vision.

% Investing a substantial effort towards the most excluded groups does not mean neglecting the other communities, but will instead ensure equity, guide the vision of diversity for your conference, and help design strategies to achieve it.
% vision
% back to the indicators? 
% realistic but bold (back?)
% be aware and promote awareness
%¡LAS BARRERAS!
%Express the diversity and inclusion vision for your conference by writing a diversity statement. 
% As a starting point, identify which groups are usually unsupported or discriminated against in the community or event of interest, and determine the contributing factors. Next, set measurable indicators of success when organizing a diverse conference. The numbers resulting from these indicators are not the primary goal, and this is not about performative allyship. Instead, these numbers should guide and help hold the organizing team accountable towards these diverse and inclusive goals.

% Rocío: more effort towards including the more excluded
% This may translate into different actions depending on your field, region, or community, \textit{e.g.}, a gender distribution of your speakers that is representative of the general population, presence of racialized people --especially Black people-- among the head organizers, speakers, and attendees, LGBTQIA+ friendly-spaces, or community participation from key geographic regions.

@@

% Rocío: main idea: Inclusion for everyone
While it is common to think about representation in terms of some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a genuinely inclusive environment means taking care of all aspects of diversity. Inclusivity should lie at the heart of setting the tone and meeting space for any meeting as we move forward in the 21st century!
% Edit and make more specific (below)

\emoji{bulb} \textbf{Practical tips:} Here are a few ways in which conference organizers can actively enforce inclusivity, such as i) Promote respect of pronouns, ii) schedule events exclusive for diverse groups (\textit{e.g.}, LGBTQIA+-friendly spaces), iii) if the conference has an in-person component, pre-assign spaces for religious practices, lactation, and child care, iv) include menu options that account for diverse dietary and religious requirements. These are just a few easy-to-implement examples with a significant impact on including key sub-groups that otherwise find it challenging (see \cite{noauthor_discover2021} for some great advice on 'low-hanging fruits').
Importantly, by being proactive in creating such a welcoming space, the organizers can be respectful of attendees' privacy and avoid requiring anyone to disclose personal information, such as revealing a disability, sexual orientation, gender identity, or mental health issues, to a name a few.

The organizers should promote a supportive and respectful environment, where all attendants can be themselves without fear of rejection or isolation. This includes promoting the idea that anyone can act in defense of other attendants in case anything unpleasant, incorrect, of offensive happens. Instead of looking away, act as "active bystanders" and intervene to keep the community safe.
[ast: should we add pre-conference allyship workshops? i saw a conference that had one pre-conference events for the whole community, just brilliant]
% Rocío: main idea: Code of conduct for a safe place
However, the most important step that organizers should take to ensure the safety of the community is
Adopting a Code of Conduct (CoC) and gathering a diverse team to enforce it \cite{favaroYourScienceConference2016}.
The CoC is a living document meant to keep the community safe and should state clearly the unacceptable behaviors, the consequences for engaging in such behavior, and the way to report violations \cite{auroraHowRespondCode2019}. 
The CoC team must receive training on how to receive reports, respond to incidents, communicate their responses, and organize protocols on how to respond correspondingly. This should be done during the early phases of the conference planning because CoC violations can happen before the conference too (\textit{e.g.}, social media, organization meetings).  
Since people from marginalized groups are more likely to be victims of Code of Conduct violations and find their claims dismissed due to inadvertent power dynamics, assembling a diverse CoC team will be essential to appropriately deal with discrimination and harassment issues and the power dynamics underneath them with respect and sensitivity. A well-drafted CoC may also be more understanding and accommodating of different disciplinary cultures and geographical and cultural considerations. 
We strongly recommend reading `How to Respond to Code of Conduct Reports' \cite{auroraHowRespondCode2019} as an excellent starting point for building a strong CoC team.

@@

A genuinely inclusive conference can only be organized by an inclusive and diverse team in charge of the conference decision-making.
Organizers should be a team with people from different regions, genders, ethnicities, socioeconomic statuses, and other aspects of diversity.
Having diverse people in decision-making roles will positively affect the conference as a whole because all the processes will benefit from their expertise, experience, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. 
In addition, a diverse team plays an essential role in creating a welcoming space because representation -- seeing people with similar life experiences occupy critical positions of power and breaking negative stereotypes -- is one of the best ways to create a sense of belonging for everyone participating in the conference (see \textbf{Rule 2} about creating a safe and welcoming environment).

% Rocío: Idea of this paragraph: non-transferability and do not restrict people
People with disabilities often say: 'Nothing about us without us' [ref]; the same holds for other dimensions of diversity. The experience of people in marginalized groups facing barriers for equal participation and trying to solve them are invaluable; they cannot be replaced by good intentions or second-hand knowledge from people who have not lived through the same experiences \cite{costanzachockDesign2020}.
% People who have experienced exclusion have the social and technical expertise to fight against it and they will be essential in any team.%ast maybe frame this on a more positive note (ideas, problem solving)
It is important, though, to ensure that organizers with diverse backgrounds are not restricted to only work—or want to work—on diversity and inclusion aspects; 
every member should be granted the freedom to choose which areas of the conference they want to work on, in addition to contributing to decisions involving marginalized communities. Needless to say, they are respected members of the scientific community, and they should be treated as such.

@@

\emoji{bulb} \textbf{Practical tips:} If you have already started assembling an organizing team, check for gaps in its composition and find representatives accordingly. 
The regional and local communities, groups, or associations in your field are good sources to tap into. 
Furthermore, self-nomination and voting do not always work as reliable mechanisms to fill leadership positions. Instead, it helps to nominate members directly and offer them leading positions in the organization team, especially those that would be typically occupied by people from privileged groups.
Build an environment that encourages healthy conflict and constructive criticism, one in which every person can express their opinion and pay special attention to people from systematically excluded groups.
Many people lack the institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free; consider paying them as an item in your budget as well as other types of recognition (\textit{e.g.}, public recognition on social media, on the conference website, or during the conference, ideally whatever works best for the contributing member).
In addition, tasks such as receiving and responding to Code of Conduct reports can be emotionally intense work and should be additionally rewarded.

@@

The next set of rules concern diverse elements in the design of the conference and operational tasks. They focus on weaving inclusion into the conference design process.
In \textbf{Rule 4}, we introduce ways to counteract bias in the conference program, keynotes, program committee, abstract selection, and other spaces. 
\textbf{Rule 5} describes the advantages of online conferences by reducing barriers for participation and provides advice for organizing online and hybrid conferences.
\textbf{Rule 6} focuses on best practices to make all conference platforms and spaces accessible to people with disabilities. 
In \textbf{Rule 7}, we advocate for more language-inclusive conferences and provide suggestions to encourage and ensure full participation of non-native English speakers. 
\textbf{Rule 8} provides tips for developing an inclusive media strategy. 
In \textbf{Rule 9}, we address budgeting for inclusive practices, and helping participants with affordable registration costs, scholarships, and other forms of financial support.

@@

When choosing or inviting people for visible and valued roles in the conference––such as keynote speakers, program/scientific committee members, session chairs––, it is likely that there will not be much diversity in the first set of names.
Our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyerNoticeWhoScience2021,swartzScienceValueDiversity2019,wongBuildDiversityScience2020,dignazioUnicornsJanitorsNinjas2020}. 
Rather than deter us, this should encourage us to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard great people that are not routinely in the spotlight, making sure that these roles do not stay with people in privileged groups (e.g. avoid `manels' and all-white panels, for instance).
Make sure the composition of all these groups is balanced across topics, areas, and representation across regions, gender, race, career stage, among other aspects.

@@

The final list of accepted talks can also be biased in both submission and acceptance processes. 
People who have never participated in conferences, and especially those regarded as big and prestigious, may not feel confident to submit their work.
To counteract this self-selection bias, promote the submission call beyond the communication channels and reach out to local groups and communities of practice to encourage submissions by their members.
Some of these communities may have pre-submission mechanisms to help their members prepare and improve their submissions (e.g. the Latin-R community has a Slack channel with the purpose of helping community members improve their abstracts for conferences, and the R-Ladies community did the same for useR! 2021).
On the other hand, the review process may be biased because selection committees—even diverse ones—can be subjected to unconscious or implicit bias.
Unconscious bias happens when we interpret information, such as names of the authors, we assume their origin or ethnicity, their institution of origin, and we attach a positive or negative value to their work according to negative or positive stereotypes. Nobody is free from implicit bias [cite Ross2019].
Selection committees need to be reminded of such biases (the implicit association test can help identifying them [cite Greenwald - Measuring individual differences in implicit cognition: The implicit association test]), and
decide on the most appropriate strategy to counteract them, either with double blind, double open, or single blind reviews [cite discover cookbook].
Double blind reviews can help counteract implicit bias only if it is possible to completely anonymize the identifying information from submission. 
Double open, though more transparent, may still be subjected to implicit bias, and reviewers from minoritized groups might feel additional pressure because the reviews are signed. If this likely case, and in fields were it is not possible to guarantee anonymity of the authors during the selection process (e.g.  non-anonymized GitHub repositories need to be examined), single blind reviews are the only option left.  
In any case, create clear evaluation criteria for the submitted proposals that do not put people from minoritized groups at disadvantage (e.g. based on their number of publications or talks), and share them with the authors and reviewers.
Advertising your practices to mitigate bias in proposal selection may also encourage people from diverse groups to submit their work, thus counteracting self selection bias.

@@

% Rocío: main idea: unbias roles and give spotlight to the roles--and people who have contributed in roles--that are not stereotypically categorized as success
Implicit bias also pertain the value that we give to certain activities; e.g. scientific publication and development of software are often regarded as higher value than other roles like community building or teaching. 
These roles may be equally if not more challenging and are usually assumed by women, people of color, people with disabilities, and other minoritized groups \cite{cheng2020x+, burfordHomelinessMeantHaving2020}. Research about these topics is undervalued relative to other areas of knowledge production. 
Counteracting bias also means innovating the program of the conference, proposing new thematic sessions, broadening the scope of talks, keynotes, and tutorials, and giving visibility to the whole range of activities and practitioners that contribute to the overall topics of the conference.

@@

% Be more direct
Conferences are among the least accessible spaces that people with disabilities may encounter in professional contexts \cite{priceAccessImaginedConstruction2009}. Even when conferences implement other inclusive practices, the participation of people with disabilities is often overlooked \cite{marks2021meeting}. This is one key aspect where participation in the organizing team allows disabled people to take part in the decisions from the beginning (see rule 3 about the organizing team), as thinking about disability or simulating it are not substitutes for real-life experience \cite{costanzachockDesign2020}. Planning for accessibility requires time and early decision-making \cite{irishIncreasingParticipationUsing2020}. As other aspects of inclusion, dealing with accessibility at the last minute is the recipe for a disastrous conference. If you do not consider accessibility from the conference inception, it is highly likely that you will be better off without trying to patch accessibility at the last minute. Key decisions in this respect are hard to correct.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}

The authors of this article recently organized useR! 2021, the annual global R conference, attracting a broad range of participants from academia, industry, government, and the non-profit sector. To embrace the constraints posed by the pandemic, we hosted a 100\% virtual conference. We worked on building a high-quality virtual and global conference by creating a kind, inclusive, accessible, and welcoming environment for everyone. 
This article outlines our most important lessons learned abstracted as `10 simple rules to host an inclusive conference'. These rules apply equally to academic, industry, or mixed conferences; while inspired by a global experience, these rules seamlessly translate to regional events. 
We have designed them for wide adoption to enable and ensure that the scientific and technical conferences are as accessible and inclusive as they can be.

@@

% Rocío: Main message of the paragraph: Conferences are great, but for some
Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions.
A good conference experience can be 
However, opportunities for participating in conferences are not equal for all. 
Many academic and tech conferences have become spaces that reproduce systemic inequalities and privileges, i.e. advantages given by society to people from some groups, like being white, male, from a rich country, English-native speaker, with no physical disabilities \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}.
% However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields.

This article proposes 10 simple rules to host inclusive conferences.
They are organized in three sections: \textbf{the pillars}, fundamental rules with key elements to conceive the work on diversity and inclusion in any conference (embracing diversity, creating a welcoming environment, and having a diverse organizing team); \textbf{design rules}, concerning diverse elements in the design of the conference and operational tasks (counteracting bias in the conference program, embracing a strong online component, achieving accessibility for people with disabilities, embracing language diversity, enforcing inclusive communication strategies, and allocating financial resources to support inclusive practices); and a \textbf{process rule} to make the conference part of a long-term commitment to inclusion. 
These practices stem from the experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}.
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many people as possible.

The authors of this article include the global coordinators of the conference, members of the diversity and inclusion team, and people who contributed to other areas of the conference. We come from different backgrounds, career stages and even continents.
The rules result from our lessons learned from diverse perspectives and expertise. 
Even though these rules were designed based on our experience with a global event, they can be easily transferred or scaled to events with regional or local scopes.
These rules will greatly benefit people who are part of meeting committees that oversee the site/location selection process, or that coordinate with the local organizers of conferences.
The rules will also help organizers who intend to put inclusion at the core of their conference right from the conception and planning phases. We have designed them for wide adoption to enable and ensure that the scientific and technical conferences are as accessible and inclusive as they can be.

@@

\begin{figure}[!h]
\centering
%sara: without success i'm trying to use pdfcomment package and pdftooltip to make the alt-text. any help?
%ast: R/ Unfortunately, support for adding accessibility metadata to LaTeX documents is limited. You will probably need to use Adobe Acrobat to add missing accessibility metadata to your PDF file.https://chi2021.acm.org/for-authors/presenting/papers/guide-to-an-accessible-submission#latex %ast: let's try not to forget this when submitting - some edits to the PDF may be needed to add alt-text to the final version
\pdftooltip{
\includegraphics[width=\textwidth]{figs/rules.png}}{
%sara: suggestion for alt text:
%Diagram of how the 10 rules are organized into three groups: grounding rules (rules 1, 2, and 10), people-related rules (rules 3 and 4), and design rules (rules 5 to 9). The diagram has five rows. Each line contains rectangular boxes with the rule number and a short title for each rule. Each rule box is colored based on their group (grounding rules are grey, people-related rules are green and design rules are pink). First row is for the grounding rules 1. Multidimensional diversity and 2. Safe and inclusive environment in grey boxes. Second row is for people-related rules 3. Organizing team and 4. Unbias in green boxes. Third row is for the design rules 5. Online component, 6. Accessibility for disabilities, 7. Language inclusivity in pink boxes. The fourth row is for the design rules 8. Communication and 9. Financial support and budget in pink boxes. The last row is for the grounding rule 10. Diversity and inclusion as a process in grey boxes.
}
\caption{Diagram of the rules organized in three groups: the pillars (rules 1, 2, and 3), design rules (rules 4 to 9), and the process (rule 10).}
\label{fig:diagram}
\end{figure}

@@

To organize a more inclusive conference, it is necessary to identify upfront which groups of people are being excluded, the barriers that keep them from active partaking, and strategies to overcome them. 
\textbf{Rule 1} is about embracing all dimensions of diversity and prioritizing the participation of groups that are often marginalized.
However, increasing diversity alone is not enough. \textbf{Rule 2} focuses on how to create a safe and welcoming environment for all the conference attendees. 
Inclusion should start from the organizing team. \textbf{Rule 3} highlights the importance of starting with an inclusive and diverse organizing team, and provides tips on work dynamics.

@@

% Rocío: Main idea: what is diversity and why does it matter? (inequalities)
Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few.
Human diversity should be celebrated and respected in every way. 
Unfortunately, we live in a world with implicit hierarchies along these axes. 
Some statuses (e.g. cisgender, white, male, from North America or Western Europe) hold the privilege of being the defaults around which all systems—including conferences—are consciously and unconsciously built. 
People outside the dominant groups suffer different forms of oppression, like sexism, racism, ableism, homophobia, and transphobia; some of them have to deal with several of these on a daily basis.
% While no isolated initiative can drastically alter xx, building a more diverse and inclusive conference starts by recognizing that these inequalities have systematically excluded whole groups of people from academia and scientific and professional circles. 
Therefore, recognizing this systematic exclusion, particularly in our fields and in our scientific or professional communities, is paramount to inclusion, and will help identify the most marginalized and discriminated subgroups \cite{timperleyHeMoanaPukepuke2020}.
% These are the groups that warrant deliberate efforts that ensure fair inclusion.

As a starting point, the meeting organizers could identify which groups warrant deliberate efforts to ensure fair inclusion, and determine the contributing factors (e.g. explicit discrimination, participation fees).
This analysis should guide the vision of diversity and inclusion for your conference and help design strategies to achieve it. 
Express this vision for your conference in a diversity statement. 
Next, use this vision and the set strategies to define measurable indicators and goals of diversity and inclusion (e.g. a gender distribution of your speakers that is representative of the general population, the presence of racialized people (e.g. Black people) among the organizers, speakers, and attendees, or participation from key geographic regions). 
These numbers should guide you and help achieve your vision.

% realistic but bold (back?)
% be aware and promote awareness
% Rocío: more effort towards including the more excluded
% This may translate into different actions depending on your field, region, or community

@@

% Rocío: main idea: Inclusion for everyone
While it is common to think about representation in terms of some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a genuinely inclusive environment means taking care of all aspects of diversity. %Inclusivity should lie at the heart of setting the tone and meeting space for any meeting as we move forward in the 21st century!
% Edit and make more specific (below)
%\emoji{bulb} \textbf{Practical tips:} 
Here are a few ways in which conference organizers can actively enforce inclusiveness, such as i) Promote respect of pronouns, ii) schedule events exclusive for diverse groups (e.g. LGBTQIA+-friendly spaces), iii) if the conference has an in-person component, pre-assign spaces for religious practices, lactation, and child care, iv) include menu options that account for diverse dietary and religious requirements. These are just a few easy-to-implement examples with a significant impact on including key sub-groups that otherwise find it challenging (see \cite{noauthor_discover2021} for some great advice on `low-hanging fruits').
Importantly, by being proactive in creating such a welcoming space, the organizers can be respectful of attendees' privacy and avoid requiring anyone to disclose personal information, such as revealing a disability, sexual orientation, gender identity, or mental health issues, to name a few.

The organizers should promote a supportive and respectful environment, where all attendants can participate without fear of rejection or isolation. This includes promoting the idea that anyone can act in defense of other attendants in case anything unpleasant, incorrect, of offensive happens. Instead of looking away, act as "active bystanders" and intervene to keep the community safe.
% Rocío: main idea: Code of conduct for a safe place %i think the transition here should leave the two concepts closer. not however. 
However, the most important step that organizers should take to ensure the safety of the community is
Adopting a Code of Conduct (CoC) and gathering a diverse team to enforce it \cite{favaroYourScienceConference2016}.
The CoC is a living document meant to keep the community safe and should state clearly the unacceptable behaviors, the consequences for engaging in such behavior, and the way to report violations \cite{auroraHowRespondCode2019}. 
The CoC team must receive training on how to receive reports, respond to incidents, communicate their responses, and organize protocols on how to respond correspondingly. This should be done during the early phases of the conference planning because CoC violations can happen before the conference too (e.g. social media, organization meetings).  
Since people from marginalized groups are more likely to be victims of Code of Conduct violations and find their claims dismissed due to inadvertent power dynamics, assembling a diverse CoC team will be essential to appropriately deal with discrimination and harassment issues and the power dynamics underneath them with respect and sensitivity. A well-drafted CoC should account for  different cultures and geographical and cultural considerations. 
We strongly recommend reading `How to Respond to Code of Conduct Reports' \cite{auroraHowRespondCode2019} as an excellent starting point for building a strong CoC team.

@@

A genuinely inclusive conference can only be organized by an inclusive and diverse team in charge of the conference decision-making.
Organizers should be a team with people from different regions, genders, ethnicities, socioeconomic statuses, and other aspects of diversity.
Having people from diverse backgrounds in decision-making roles will positively affect the conference as a whole because all the processes will benefit from their expertise, experience, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. 
In addition, a diverse team plays an essential role in creating a welcoming space because representation—seeing people with similar life experiences occupy critical positions of power and breaking negative stereotypes—is one of the best ways to create a sense of belonging for everyone participating in the conference (see \textbf{Rule 2} about creating a safe and welcoming environment).

% Rocío: Idea of this paragraph: non-transferability and do not restrict people
People with disabilities often say: `Nothing about us without us' [ref]; the same holds for other dimensions of diversity. The experience of people in marginalized groups facing barriers for equal participation and trying to solve them are invaluable; they cannot be replaced by good intentions or second-hand knowledge from people who have not lived through the same experiences \cite{costanzachockDesign2020}.
% People who have experienced exclusion have the social and technical expertise to fight against it and they will be essential in any team.%ast maybe frame this on a more positive note (ideas, problem solving)
It is important, though, to ensure that organizers with diverse backgrounds are not restricted to only work on diversity and inclusion aspects; 
every member should be granted the freedom to choose which areas of the conference they want to work on, in addition to contributing to decisions involving marginalized communities.

@@

%\emoji{bulb} \textbf{Practical tips:} 
If you have already started assembling an organizing team, check for gaps in its composition and find representatives accordingly. 
The regional and local communities, groups, or associations in your field are good sources to tap into. 
Furthermore, self-nomination and voting do not always work as reliable mechanisms to fill leadership positions. Instead, it helps to nominate members directly and offer them leading positions in the organization team, especially those that would be typically occupied by people from privileged groups.
Build an environment that encourages healthy conflict and constructive criticism, one in which every person can express their opinion and pay special attention to people from systematically excluded groups.
Many people lack the institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free; consider paying them as an item in your budget as well as other types of recognition (e.g. public recognition on social media, on the conference website, or during the conference, ideally whatever works best for the contributing member).
Furthermore, tasks such as receiving and responding to Code of Conduct reports can be emotionally intense work and should be additionally rewarded.

@@

The next set of rules concern diverse elements in the design of the conference and operational tasks. They focus on weaving inclusion into the conference design process.
In \textbf{Rule 4}, we introduce ways to counteract bias in the conference program (keynotes, program committee, abstract selection, and other spaces). 
\textbf{Rule 5} describes the advantages of online conferences by reducing barriers for participation and provides advice for organizing online and hybrid conferences.
\textbf{Rule 6} focuses on best practices to make all conference platforms and spaces accessible to people with disabilities. 
In \textbf{Rule 7}, we advocate for more language-inclusive conferences and provide suggestions to encourage and ensure full participation of non-native English speakers. 
\textbf{Rule 8} provides tips for developing an inclusive media strategy. 
In \textbf{Rule 9}, we address budgeting for inclusive practices, and helping participants with affordable registration costs, scholarships, and other forms of financial support.

@@

When choosing or inviting people for visible and valued roles in the conference –– such as keynote speakers, program/scientific committee, session chairs –– it is likely that there will not be much diversity in the first pool of names.
This is likely due to our inherent biases resulting from existing systems that have always privileged some groups of people \cite{dwyerNoticeWhoScience2021,swartzScienceValueDiversity2019,wongBuildDiversityScience2020,dignazioUnicornsJanitorsNinjas2020}. 
Rather than deter us, this should encourage us to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard great people that are not routinely in the spotlight, making sure that these roles do not stay with people in privileged groups (e.g. avoid `manels' and all-white panels; \cite{else_how_2019}; see \textbf{Rule 8}).
\emoji{bulb} It would help to ensure that the composition of all these groups is balanced across topics and areas with representation across regions, gender, race, and career stage, among other aspects.

@@

The final list of abstracts making it into the program could also be biased in both submission and acceptance processes. 
For instance, first-time conference participants, especially for the flagship/prestigious meetings, may not feel confident to submit their work.
\emoji{bulb} To counteract this self-selection bias, promote the submission call beyond the usual communication channels and reach out to local groups and communities of practice to encourage submissions by their members.
Some of these communities may have pre-submission mechanisms to help their members receive feedback for their submissions (e.g. the Latin-R and R-Ladies communities have Slack channels to help their members improve their conference abstracts).
On the other hand, the review process may be biased because selection committees — even diverse ones — can be subjected to unconscious or implicit bias.
Unconscious bias happens when we interpret information such as names of the authors, assuming their origin, ethnicity, or their institution of origin, and we attach a positive or negative value to their work according to the stereotypes. Unfortunately, nobody is free from implicit bias \cite{ross_everyday_2020}.
Selection committees need to be gently reminded of such biases with evaluation guidelines (the implicit association test can help identifying them \cite{greenwald_measuring_nodate}) and the organizers need to
identify the most appropriate strategy to counteract them, either with double blind, double open, or single blind reviews \cite{numfocus_discover_2021}.

Double blind reviews can help counteract implicit bias only if the identifying information is completely anonymized in the submission. 
Double open reviews, though more transparent, may still be subjected to implicit bias, and reviewers from minoritized groups might feel additional pressure because the reviews are signed. In this likely case, and in fields where it is not possible to guarantee anonymity of the authors during the selection process (e.g. when non-anonymized GitHub repositories need to be examined), single blind reviews are the only remaining option.

%\emoji{bulb} \textbf{Practical tips:} 
In any case, design unambiguous evaluation criteria for the submitted proposals that do not put people from minoritized groups at disadvantage (e.g. based on their number of publications or talks), and share them with the authors and reviewers.
Advertising your practices to mitigate bias in proposal selection may also encourage people from diverse groups to submit their work, thus counteracting self-selection bias.

@@

% Rocío: main idea: unbias roles and give spotlight to the roles--and people who have contributed in roles--that are not stereotypically categorized as success
Implicit bias also manifests in the weightage given to certain products/activities; e.g. scientific publication and development of software are often rated higher than community building or teaching. 
The latter roles are equally, if not more, challenging and are usually assumed by women, people of color, people with disabilities, and other minoritized groups \cite{cheng2020x+, burfordHomelinessMeantHaving2020}. 
Likewise, research about these topics is undervalued relative to other areas of knowledge production. 
Counteracting bias actively means innovating the program of the conference, proposing new thematic sessions, broadening the scope of talks, keynotes, and tutorials, and giving visibility to the whole range of activities and practitioners that contribute to the overall range and quality of the conference.

@@

% Be more direct
Conferences are among the least accessible spaces that people with disabilities may encounter in professional contexts \cite{priceAccessImaginedConstruction2009}. Even when conferences implement other inclusive practices, the participation of people with disabilities is often overlooked \cite{marks2021meeting}. Including people with disabilities in the organizing team can have a large impact from the beginning (see \textbf{Rule 3}), since planning for accessibility requires time, experience, and early decision-making \cite{irishIncreasingParticipationUsing2020}. As other aspects of inclusion, dealing with accessibility at the last minute is the recipe for a disastrous conference. If you do not consider accessibility from the conference inception, it is highly likely that you will be better off without trying to patch accessibility at the last minute. Key decisions in this respect are hard to correct.

If the conference has an in-person component, the venue should comply with common accessibility standards, such as being adequate for people who use wheelchairs, have signs in Braille, and a sound system compatible with hearing devices and live interpretation, just to name a few important features. In addition to this, the organizers should take care proactively of invisible disabilities (e.g. dyslexia, anxiety, ADHD, autism). This could be done in many ways, for example, by providing quiet spaces for privacy and noise-free conversations, or providing chairs in open spaces.
